---
title: Height and NCAA Hoops
author: ''
date: '2017-12-21'
slug: height-and-ncaa-hoops
categories:
  - R
  - NCAA
tags:
  - rstats
  - basketball
  - ncaa
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(collapse = TRUE, warning = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)
library(rvest)
library(magrittr)
library(knitr)
library(ggthemes)
```

```{r kp_get_height, eval = TRUE, echo = FALSE, message=FALSE, warning=FALSE}
#Data scraped using rvest, but including source file here in csv format since the script required authentication details, in a private repo.
kp_height_exp <- read.csv("kp_height_exp.csv")

```

I'm hellbent on winning a March Madness pool this year, and have been reading up on modern basketball analytics (i.e., [KenPom](http://kenpom.com), [Dean Oliver](http://www.basketballonpaper.com/author.html), [Sagarin](http://sagarin.com/sports/cbsend.htm), [Lopez](https://statsbylopez.com/2014/12/04/building-an-ncaa-mens-basketball-prediction-model/)).

In the process, I've played around with scraping data from various websites, to analyze in R.

This is a quick exploration of player height according to 2017-18 NCAA D-I Men's Basketball Rosters.

***
### Average Height by Team
```{r team_height, echo = FALSE, eval = TRUE}
kp_height_exp %>% 
  filter(!Team == "Team") %>% 
  ggplot(aes(x = Avg.Hgt))+
  geom_density()+
  theme_fivethirtyeight()+
  scale_x_continuous(name = "Average Height (inches)",
                              breaks = seq(74, 80, 1))+
  ylab("Density")

```

D-I teams tend to be around `6'5"` in average height -- 8 inches taller than the [average US adult male](https://www.cdc.gov/nchs/fastats/body-measurements.htm) at `5'9"`.

### 10 Tallest Conferences

Averaging the heights of players on each team's roster:

```{r tall_ht_by_conf, echo = FALSE, eval = TRUE}
kp_height_exp %>% 
  filter(!Team == "Team") %>% 
  group_by(Conf) %>% 
  summarise("Average height" = paste0(mean(`Avg.Hgt`)%/%12,"'",round(mean(`Avg.Hgt`)%%12,2),"''")) %>% 
  arrange(desc(`Average height`)) %>% 
  head(10) %>% 
  kable()

```

There are `33` NCAA D-1 Conferences, and all of the perennial top-performers are among the tallest. 


### 10 Shortest Conferences


```{r short_ht_by_conf, echo = FALSE, eval = TRUE}
kp_height_exp %>% 
  filter(!Team == "Team") %>% 
  group_by(Conf) %>% 
  summarise("Average height" = paste0(mean(`Avg.Hgt`)%/%12,"'",round(mean(`Avg.Hgt`)%%12,2),"''")) %>% 
  arrange(`Average height`) %>% 
  head(10) %>% 
  kable()

```

At the time of this writing, not a single team from what I'll now call the `Short 10 Consortium` is ranked in the Top 60 according to [Kenpom](http://kenpom.com), nor ESPN's [BPI](http://www.espn.com/mens-college-basketball/bpi/_/view/bpi).

```{r rebounding_data, echo=FALSE, eval = TRUE}
### Add kp_ids
kp_ids <- read.csv("kp_team_ids.csv")

kp_height_exp <- kp_height_exp %>% 
  left_join(kp_ids, by = c("Team"="team"))

# Load SR Data

team_stats <- read_html("https://www.sports-reference.com/cbb/seasons/2018-school-stats.html") %>% 
  html_nodes("#basic_school_stats") %>% 
  html_table(fill = TRUE) %>% 
  as.data.frame()

# Add colum headers, filter out extra rows
headers <- c("Rank","School","G","W","L","WPct","SRS","SOS","W_conf","L_conf",
             "W_home","L_home","W_away","L_away","Pts","PtsA","break","MP","FG","FGA",
             "FG_pct","3P","3PA","3P_pct","FT","FTA","FT_pct","ORB","TRB",
             "AST","STL","BLK","TOV","PF")

colnames(team_stats) <- headers
rm(headers)

team_stats <- team_stats %>% 
  filter(!Rank %in% c("Rk","")) %>% 
  select(-`break`)


# Update class to numeric
team_stats %<>% mutate_at(c(1,3:33), funs(as.numeric(.)))


#############################
## Get Advanced Stats from SR
#############################

## Get team advanced stats from sr from kenpom
team_adv_stats <- read_html("https://www.sports-reference.com/cbb/seasons/2018-advanced-school-stats.html") %>% 
  html_nodes("#adv_school_stats") %>% 
  html_table(fill = TRUE) %>% 
  as.data.frame()

# Add colum headers, filter out extra rows
headers <- c("Rank","School","G","W","L","WPct","SRS","SOS","W_conf","L_conf",
             "W_home","L_home","W_away","L_away","Pts","PtsA","break", "Pace",
             "ORtg", "FTr", "3PAr", "TS_pct", "TRB_pct", "AST_pct", "STL_pct", 
             "BLK_pct", "eFG_pct", "TOV_pct", "ORB_pct", "FT_FGA")

colnames(team_adv_stats) <- headers
rm(headers)

team_adv_stats <- team_adv_stats %>% 
  filter(!Rank %in% c("Rk","")) %>% 
  select("School","Pace","ORtg", "FTr", "3PAr", "TS_pct", "TRB_pct",
         "AST_pct", "STL_pct", "BLK_pct", "eFG_pct", "TOV_pct",
         "ORB_pct", "FT_FGA")

# Update class to numeric
team_adv_stats %<>% mutate_at(c(2:14), funs(as.numeric(.)))

# Join advanced stats to the main team_stats table
team_stats <- team_stats %>% 
  left_join(team_adv_stats, by = "School")

rm(team_adv_stats)

# Add team IDs to SR data
sr_ids <- read.csv("sr_team_ids.csv")

team_stats <- team_stats %>% 
  left_join(sr_ids, by = "School")

# Join rebounding data from SR Team Data
rebounds <- team_stats %>% 
  select(team_id, ORB_pct, BLK_pct, TRB_pct)

kp_height_exp <- kp_height_exp %>% 
  filter(!Team == "Team") %>% 
  left_join(rebounds, by = "team_id")

# Clean up unneeded objects in workspace
rm(list = c("kp_ids","sr_ids", "team_stats"))

```

### Do taller teams tend to rebound, and block shots more than shorter teams?

```{r off_rebounding_corr, eval = TRUE, echo = FALSE}

ggplot(kp_height_exp, aes(x=Avg.Hgt, y=ORB_pct))+
  geom_point()+
  ggtitle("Team Height vs. Off. Rebounding Pct")
```
Correlation coefficient: `r round(cor(kp_height_exp$Avg.Hgt, kp_height_exp$ORB_pct),3)`


```{r tot_rebounding_corr, eval = TRUE, echo = FALSE}
ggplot(kp_height_exp, aes(x=Avg.Hgt, y=TRB_pct))+
  geom_point()+
  ggtitle("Team Height vs. Total Rebounding Pct")
```
Correlation coefficient: `r round(cor(kp_height_exp$Avg.Hgt, kp_height_exp$TRB_pct),3)`


```{r block_pct_corr, eval = TRUE, echo = FALSE}
ggplot(kp_height_exp, aes(x=Avg.Hgt, y=BLK_pct))+
  geom_point()+
  ggtitle("Team Height vs. Block Pct")
```
Correlation coefficient: `r round(cor(kp_height_exp$Avg.Hgt, kp_height_exp$BLK_pct),3)`

Looks like there is a correlation between team height, rebounding, and especially block percentage. The particularly-tall, high-block percentage team is one of my favorites: `Syracuse.`

***

Thank you to [Ken Pomeroy](kenpom.com) and [Sports-Reference](sports-reference.com) for their stats.
