---
title: 'What''s the plural of Eephus?'
subtitle: '...and other questions about MLB pitch velocity' 
output: html_document
date: "2017-11-11"
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(collapse = TRUE, cache = TRUE, warning = FALSE)
library(dplyr)
library(ggplot2)
```

```{r get_pitches, eval = TRUE, echo = FALSE}

pitch_data <- read.csv("pitch_data_prod.csv")

```

***

I downloaded a fairly large amount of pitch data from the [MLB Statcast](https://baseballsavant.mlb.com/statcast_search) database: every pitch thrown during the 2017 season, with ~75 variables about each.

### What's the plural of Eephus?

For the uninitiated, [the Eephus](https://en.wikipedia.org/wiki/Eephus_pitch) pitch is a very, very slow baseball pitch. I'd hoped that the plural of eephus would be eephucies (like matrices), but it turns out isn't a real word at all. As far as I can tell from the [Google Ngram](https://books.google.com/ngrams/graph?content=eephus&year_start=1800&year_end=2000&corpus=15&smoothing=3&share=&direct_url=t1%3B%2Ceephus%3B%2Cc0#t1%3B%2Ceephus%3B%2Cc1) viewer, it was first used in a baseball context around 1960. More on eephuses/eephucies later.

### Which teams threw the hardest on average?

```{r team_averages, echo = FALSE, eval = TRUE}
pitch_data %>%
  group_by(mlb_team) %>%
  summarize("Release Speed" = mean(release_speed)) %>%
  arrange(desc(`Release Speed`)) %>%
  mutate("rank" = rank(-`Release Speed`))

```

Eight of the top ten hardest throwing teams are in the National League, but since these rankings are likely buoyed by the relative velocity of fastballs thrown, I wanted to re-run the list by `fastballs` alone.

### Which teams threw the fastest fastballs? (Part 1)
```{r team_averages_fastballs, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type %in% c("FF","FT", "SI")) %>%
  group_by(mlb_team) %>%
  summarize("Release Speed" = mean(release_speed)) %>%
  arrange(desc(`Release Speed`)) %>%
    mutate("fastball rank" = rank(-`Release Speed`)) 

```

Since there was a bit of variance between the overall velocity and fastball velocity rankings, I wanted to get a sense of the MLB-wide velocity by `pitch type`. I didn't remove the `Null` or `unknown` pitch types from the dataset, but I am curious as to how `AJ Ramos` chucked `11` nulls during the 2017 season.

### Average velocity by pitch type
```{r pitch_averages, echo = FALSE, eval = TRUE}
pitch_data %>%
  group_by(pitch_type) %>%
  summarize("Release Speed" = mean(release_speed)) %>%
  arrange(desc(`Release Speed`)) %>%
  mutate(pitch_type = recode(pitch_type,
                             FF = "Four-Seam Fastball",
                             FT = "Two-Seam Faseball",
                             SI = "Sinker",
                             FC = "Cutter",
                             PO = "Pitchout",
                             FO = "Forkball",
                             null = "Null",
                             FS = "Splitter",
                             SL = "Slider",
                             CH = "Changeup",
                             KC = "Knuckle-Curve",
                             SC = "Screwball",
                             CU = "Curveball",
                             IN = "Intentional Ball",
                             KN = "Knuckleball",
                             AB = "Automatic Ball",
                             EP = "Eephus",
                             UN = "Unknown"))

```

Back to fastball velocity, I created a [Cleveland plot](https://www.jstor.org/stable/2288400) to visualize the aggregated averages of Four-Seam Fastball, Two-Seam Fastball, and Sinker, by team.

### Which teams threw the fastest fastballs? (Part 2)
``` {r average_fastball_by_team_cleveland, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type %in% c("FT", "FF", "SI")) %>%
  group_by(mlb_team) %>%
  summarize("avg_fb" = mean(release_speed)) %>%
  ggplot(aes(x = avg_fb, y = reorder(mlb_team, avg_fb))) +
  geom_point(size=3) + 
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour="grey60", linetype="dashed")) +
  xlab("Average Fastball Velocity") +
  ylab("Team")
```

### Pitch velocity by team (by common pitch type)
Since the cleveland plot is a nice way to compare pitch velocity by team, I added a couple of additional common pitch types:

* `Four-Seam Fastball`
* `Slider`
* `Change-up`
* `Curveball`

``` {r avg_velocity_by_pitch_type, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type %in% c("CU", "FF", "SL", "CH")) %>%
  group_by(mlb_team, pitch_type) %>%
  summarize("avg_velo" = mean(release_speed)) %>%
  ggplot(aes(x = avg_velo, y = mlb_team, col = pitch_type)) +
  geom_point(size=3) + 
  theme_light() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour="grey60", linetype="dashed")) +
  scale_color_discrete(name = "Pitch Type", breaks = c("CH", "CU", "FF", "SL"), labels = c("Change-Up", "Curveball", "Four-Seam FB", "Slider"))+
  xlab("Average release velolcity (mph)")+
  ylab("Team") +
  ggtitle("Average velocity of four pitch types")
```

Before turning my attention back to eephuses, I wondered how fastball velocities varied `by inning`.


### Fastball velocity by Inning
``` {r average_fb_velo_by_inning, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type %in% c("FF", "FT", "SI")) %>%
  group_by(inning) %>%
  summarize("Average Velocity (mph)" = mean(release_speed)) %>%
  ggplot(aes(x = inning, y = `Average Velocity (mph)`)) +
  geom_point(size = 3) + 
  theme_light() +
  theme(panel.grid.major.y = element_line(colour="grey60", linetype="dashed")) +
  scale_x_continuous(breaks = 1:12, limits = c(1,12)) +
  scale_y_continuous(breaks = 91:97, limits = c(92,95)) +
  ggtitle("Average Fastball velo by inning") +
  xlab("Inning")
```

Looks like the starting pitcher's velocity tends to fall off as they pitch deeper into a game. As relief pitchers enter the game, the average fastball velocities shoot up. Just a hunch: the velocities between the `5th` and `8th innings` are equally affected by each game's innings pitched by starting pitchers and the timing of the *typically* hard-throwing relief pitchers entering the game. In other words, I'd predict that the average starting pitcher's `IP` is around `5 innings`, which is when the average fastball velocity starts to increase.

## Eephuses/Eephucies

There were `147` eephuses thrown during the 2017 regular season.

### Who threw them?

```{r eephus_throwers, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type == "EP") %>%
  group_by(player_name) %>%
  tally() %>%
  arrange(desc(n))
```

`RA Dickey` makes sense, being a knuckleballer, but I was surprised by James Shields. The [Chicago Sun-Times](https://chicago.suntimes.com/sports/slower-is-better-for-the-reinvented-james-shields/) wrote about Shields' `curveball` registering as an `eephus` early in the season, so these pitches may have been miscoded by `Statcast`.


### Were they `balls`, `strikes`, or put `in-play`?

```{r eephus_ball_strikes_inplay, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type == "EP") %>%
  group_by(type) %>%
  tally() %>%
  arrange(desc(n))
```
The `x` denotes an `eephus` put in play by the batter... `16%` of eephuses were batted into play. 

### What were the outcomes of eephuses put in-play?

```{r eephus_outcomes, echo = FALSE, eval = TRUE}
pitch_data %>%
  filter(pitch_type == "EP" & type == "X") %>%
  group_by(description) %>%
  tally() %>%
  arrange(desc(n))
```

Shout-out to `Jose Peraza` for hitting the 2017 season's only `HR` off of an Eephus. 

***

