---
title: "Parking Violations in DC"
author: ''
date: '2018-06-29'
slug: parking-violations-in-dc
tags: []
categories: []
---

```{r setup, cache=TRUE, eval = TRUE}

```

I have very few complaints about living in Washington, D.C., but if I were forced to pick one, it'd be searching for on-street parking. When I found the several public [Parking Violations](http://opendata.dc.gov/datasets?q=Parking%20Violations&sort=-updatedAt) on [Open Data DC](http://opendata.dc.gov/), I figured I might be able to extract some insights into the best (and worst) blocks to look for a spot during high-demand times.

My original intent was to try to model the parking violations to identify the optimal times and locations of available parking spots near my house. I found the `ISSUE_TIME` data to be suspicious, and settled on doing a quick exploratory analysis of the data instead.

```{r load data, echo = FALSE, message=FALSE}
library(dplyr)
library(readr)
library(purrr)
library(lubridate)
library(ggplot2)
library(stringr)
library(tidytext)
library(knitr)
library(scales)

jun2017 <- read_csv("Parking_Violations_Issued_in_June_2017.csv")
jul2017 <- read_csv("Parking_Violations_Issued_in_July_2017.csv")
aug2017 <- read_csv("Parking_Violations_Issued_in_August_2017.csv")
sep2017 <- read_csv("Parking_Violations_Issued_in_September_2017.csv")
oct2017 <- read_csv("Parking_Violations_Issued_in_October_2017.csv")
nov2017 <- read_csv("Parking_Violations_Issued_in_November_2017.csv")
dec2017 <- read_csv("Parking_Violations_Issued_in_December_2017.csv")
jan2018 <- read_csv("Parking_Violations_Issued_in_January_2018.csv")
feb2018 <- read_csv("Parking_Violations_Issued_in_February_2018.csv")
mar2018 <- read_csv("Parking_Violations_Issued_in_March_2018.csv")
apr2018 <- read_csv("Parking_Violations_Issued_in_April_2018.csv")
may2018 <- read_csv("Parking_Violations_Issued_in_May_2018.csv")


violations <- bind_rows(jun2017, jul2017, aug2017, sep2017, oct2017, nov2017, dec2017, jan2018, feb2018, mar2018, apr2018, may2018)

rm(list = c("jun2017","jul2017","aug2017","sep2017","oct2017","nov2017", "dec2017", "jan2018", "feb2018", "mar2018", "apr2018", "may2018"))

violations <- violations %>%
  mutate(
    "day_of_week" = case_when(
      wday(TICKET_ISSUE_DATE) == 1 ~ "Sunday",
      wday(TICKET_ISSUE_DATE) == 2 ~ "Monday",
      wday(TICKET_ISSUE_DATE) == 3 ~ "Tuesday",
      wday(TICKET_ISSUE_DATE) == 4 ~ "Wednesday",
      wday(TICKET_ISSUE_DATE) == 5 ~ "Thursday",
      wday(TICKET_ISSUE_DATE) == 6 ~ "Friday",
      wday(TICKET_ISSUE_DATE) == 7 ~ "Saturday"
    )
  )



violations$day_of_week <- factor(
  violations$day_of_week,
  levels = c(
    "Sunday",
    "Monday",
    "Tuesday",
    "Wednesday",
    "Thursday",
    "Friday",
    "Saturday"
  )
)
```

------

DC Department of Transportation [(DDOT)](http://ddot.dc.gov/) publishes Parking Violations by Month. I downloaded the past year of monthly datasets and loaded them into R:


Exploring the structure of the dataset with  `summary`:
```{r explore_the_summary, echo = TRUE}
summary(violations)
```

#### Exploring missing values:  
Specifically, violation descriptions, violation locations, and violation time/date:

```{r missing_values, echo = FALSE}
violations %>% 
  summarise("Missing Descriptions" = comma(sum(is.na(VIOLATION_DESCRIPTION))),
            "Missing Locations" = sum(is.na(LOCATION)),
            "Missing Date" = sum(is.na(TICKET_ISSUE_DATE)),
            "Total Observations" = comma(n())) %>% 
  kable()
```


Slightly more than 3% of violations have no `VIOLATION_DESCRIPTION`, however, they are each reported with a `VIOLATION CODE`, summarized below:  

```{r missing_value_codes, echo = FALSE}
violations %>% filter(is.na(VIOLATION_DESCRIPTION)) %>% group_by(VIOLATION_CODE) %>% summarise("Violations" = n()) %>% arrange(desc(Violations)) %>%
mutate("Violations" = comma(Violations)) %>% 
  kable()
```

It seems like `P076` is a "[ROSA Warning,](https://dmv.dc.gov/sites/default/files/dc/sites/dmv/publication/attachments/AdjudicationDispositionRptJan2011.pdf)" which stands for Registration of Out-of-State Automobiles. Some of the other codes were hard to find, although P187 maps to... "ROLLERBLADEINFREEDOM." [Soul Skaters?](https://www.imdb.com/title/tt0162212/) 

For the rest of the exploration, I've replaced the missing `P076` Violation Description values with `ROSA WARNING`. 
```{r add_rosa_warning, echo = FALSE}
violations <- violations %>% 
  mutate(VIOLATION_DESCRIPTION = case_when(VIOLATION_CODE == "P076" ~ "ROSA WARNING", TRUE ~ VIOLATION_DESCRIPTION))
```

### Top 10 violations by description:  
```{r top_10_description, echo = FALSE}
violations %>% 
  filter(!is.na(VIOLATION_DESCRIPTION)) %>% 
  group_by(VIOLATION_DESCRIPTION) %>%
  summarise("Violations" = n()) %>% 
  arrange(desc(Violations)) %>% 
  top_n(10, Violations) %>% 
  ggplot(aes(x=Violations, y = reorder(VIOLATION_DESCRIPTION, Violations))) +
  geom_point()+
  scale_y_discrete(labels = function(x) str_wrap(x, width = 15))+
  ylab("")
```

#### Parking Violations by Month
```{r violations_by_month, echo = FALSE}
   violations %>% 
    group_by(month(TICKET_ISSUE_DATE)) %>%
    summarise("Violations" = n()) %>% 
    rename("Month" = `month(TICKET_ISSUE_DATE)`) %>% 
    mutate("Month" = month.abb[Month])  %>% 
    ggplot(aes(x=factor(Month,levels = c("Jun","Jul","Aug","Sep","Oct","Nov","Dec","Jan","Feb","Mar","Apr", "May")), y=Violations))+
    geom_bar(stat="identity")+
    xlab("")+
    ggtitle("")+
    scale_y_continuous(breaks=seq(0,125000,30000))
  
```

#### Top 10 Violation Locations
```{r top_10_violation_locations, echo = FALSE}
violations %>% 
  group_by(LOCATION) %>% 
  summarise("Violations" = n()) %>% 
  arrange(desc(Violations)) %>% 
  top_n(10, Violations) %>% 
  kable()
```

#### Violations by Day of the Week
```{r violations_by_day_of_week, echo=FALSE}
# Violations by Day of Week
violations  %>% 
  group_by(day_of_week) %>% 
  summarise("Violations" = n()) %>% 
  ungroup() %>% 
  ggplot(aes(x=day_of_week, y = Violations)) +
  geom_bar(stat="identity")+
  scale_y_continuous(limits = c(0,300000), labels = c(0,100000,20000,30000)) +
  xlab("")
```

#### Most Common Violation on each Day of the Week
```{r top_violations_by_day_of_week, echo = FALSE}

violations %>% 
  mutate("day_of_week" = case_when(wday(TICKET_ISSUE_DATE) == 1 ~ "Sunday",
                                   wday(TICKET_ISSUE_DATE) == 2 ~ "Monday",
                                   wday(TICKET_ISSUE_DATE) == 3 ~ "Tuesday",
                                   wday(TICKET_ISSUE_DATE) == 4 ~ "Wednesday",
                                   wday(TICKET_ISSUE_DATE) == 5 ~ "Thursday",
                                   wday(TICKET_ISSUE_DATE) == 6 ~ "Friday",
                                   wday(TICKET_ISSUE_DATE) == 7 ~ "Saturday")) %>% 
  group_by(day_of_week, VIOLATION_DESCRIPTION) %>% 
  summarise("Violations" = n()) %>% 
  arrange(desc(Violations)) %>% 
  group_by(day_of_week) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select("Day" = day_of_week,
         VIOLATION_DESCRIPTION) %>% 
  kable()
 
```

#### Violations by License Plate State
Top 10, in descending order 
```{r violations by state, warning=FALSE, echo = FALSE}
violations %>% 
  group_by(RP_PLATE_STATE) %>% 
  summarise("Violations" = n()) %>% 
  top_n(10, Violations) %>% 
  ungroup() %>% 
  filter(!is.na(RP_PLATE_STATE)) %>% 
  ggplot(aes(x=reorder(RP_PLATE_STATE, Violations), y = Violations)) +
  geom_point(stat="identity")+
  coord_flip() +
  scale_y_continuous(labels = comma, breaks = seq(0,500000,50000)) +
  theme(axis.text.x = element_text(angle=60, hjust=1),
        panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(colour="grey60", linetype="dashed"))+
  xlab("")+
  ylab("")+
  ggtitle("")

```

```{r violations by state table, echo = FALSE}
violations %>% 
  group_by(RP_PLATE_STATE) %>% 
  summarise("Violations" = n()) %>% 
  arrange(desc(Violations)) %>%
  ungroup() %>%  
  mutate("Percentage" = percent(Violations/sum(Violations))) %>% 
  rename("State" = RP_PLATE_STATE) %>% 
  mutate("Violations" = comma(Violations)) %>% 
  select(State,Violations,Percentage) %>% 
  head(10) %>% 
  kable()
```

[Crabcakes and parking violations](https://www.imdb.com/title/tt0396269/quotes/qt0329784), that's what Maryland does.
