---
title: "Catching up on the Tidyverse"
author: "Seth Goldman"
date: '2020-10-03'
slug: catching-up-on-the-tidyverse
categories: []
tags: []
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE) 

```

```{r load_pkgs}

library(dplyr)
library(knitr)
library(kableExtra)
```



When I first started learning R a few years ago, I focused on picking up the fundamentals of Base R, as well as the [Tidyverse](https://tidyverse.org) set of packages, concurrently. Early on, I made a point of writing as much of my code within the Tidyverse framework as possible, favoring the readability and supportive online community. 

More recently, I've realized that I haven't kept up with many of the enhancements that the Tidyverse and RStudio team have released. The two types of functionality I'd like to catch up on via this post are `tidyselect` and dplyr's `across()`.

## Tidyselect  
Tidy selection refers to the the tidy evaluation method of choosing variables, and is used in functions like `select()`, `rename()`, `pull()`, and `across()`. There's a great primer on *tidy evaluation* in the [`vignette("programming")`](https://dplyr.tidyverse.org/articles/programming.html).

Many functions that use tidy selection will include a `<tidy-select>` parameter, which means that inputs must be provided in the **tidyselect specification**.  

Using the `dplyr`'s [`tidy-select reference`](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html), I'll test some of this functionality on `dplyr`'s built-in *storms* dataset.

##### The Storms Dataset  

Sample of 5 rows:

```{r}

storm_sample <- storms %>% 
  sample_n(5) 

```

```{r tbl_storm_sample}

storm_sample %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```


### Selection Helpers

### `everything()`  
Select all variables.  


```{r everything, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(everything()) 

```
```{r tbl_everything}

storm_sample %>% 
  select(everything()) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)


```


### `last_col()`  
Select variable that matches the last column.


```{r last_col, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(last_col())

```

```{r tbl_last_col}

storm_sample %>% 
  select(last_col()) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

### Using `last_col()`, with an offset  
Select the last column, with an offset by position (i.e., the three positions from last column -- the fourth from last).

```{r last_col_offset, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(last_col(offset = 3))
```


```{r tbl_last_col_with_offset}

storm_sample %>% 
  select(last_col(offset = 3)) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

## Selection with Pattern Matching  

### starts_with()  
Select variables that start with a given prefix.  

```{r starts_with, echo=TRUE, eval = FALSE}

storm_sample %>% 
  select(starts_with("l"))

```

```{r tbl_starts_with}

storm_sample %>% 
  select(starts_with("l")) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

### ends_with()  
Select variables that end with a given suffix.  

```{r ends_with, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(ends_with("r"))

```

```{r tbl_ends_with}

storm_sample %>% 
  select(ends_with("r")) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```


### contains()  
Select varibles that contain a literal string.  


```{r contains, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(contains("sure"))

```

```{r tbl_contains}

storm_sample %>% 
  select(contains("sure")) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

### matches()  
Select variables that match a regular expression.  

```{r matches, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(matches("[aeiou]"))

```

```{r tbl_matches}

storm_sample %>% 
  select(matches("[aeiou]")) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```


### num_range()  
Select variables that match a numerical range, like x01, x02, x03.


```{r num_range, echo = TRUE, eval = FALSE}

storm_sample %>% 
  mutate("some_string1" = 1,
         "some_string2" = 0) %>% 
  select(num_range(prefix = "some_string",
                   range = 1:5))

```

```{r tbl_num_range}

storm_sample %>% 
  mutate("some_string1" = 1,
         "some_string2" = 0) %>% 
  select(num_range(prefix = "some_string",
                   range = 1:5)) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```



## Selection using character vectors  

### all_of()  
Matches variable names in a character vector. All names must be present, otherwise an out-of-bounds error is thrown.

```{r all_of, echo = TRUE, eval = FALSE}

# Assign a vector of arbitrary variables to vrbls
vrbls <- c("year","wind","status")

storm_sample %>% 
  select(all_of(vrbls))

```

```{r tbl_all_of}

# Assign a vector of arbitrary variables to vrbls
vrbls <- c("year","wind","status")

storm_sample %>% 
  select(all_of(vrbls)) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```


### any_of()  
Same as all_of(), except that no error is thrown for names that don't exist.

```{r any_of, echo = TRUE, eval = FALSE}

# Add a variable outside the scope of the dataset
vrbls2 <- c(vrbls, "dog")

storm_sample %>% 
  select(any_of(vrbls2))



```

```{r tbl_any_of}

# Assign a vector of arbitrary variables to vrbls
vrbls <- c("year","wind","status")

# Add a variable outside the scope of the dataset
vrbls2 <- c(vrbls, "dog")


storm_sample %>% 
  select(any_of(vrbls2)) %>%  
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

## Selection of variables within a function  

### where()  
Applies a function to all variables and selects those for which the function returns TRUE.

Common use case: selecting columns of a specific data type
```{r where, echo = TRUE, eval = FALSE}

storm_sample %>% 
  select(where(is.factor))

```

```{r tbl_where}
 
storm_sample %>% 
  select(where(is.factor))  %>%  
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```






## Embracing (i.e., {{var}})  
Using `{{ double curly brace }}` syntax ([pronounced 'curly curly'](https://www.tidyverse.org/blog/2019/06/rlang-0-4-0/#a-simpler-interpolation-pattern-with)) allows users to provide a variable that can be passed into a function that uses data masking or tidy selection. This helps make functions much more extensible to many types of inputs.

```{r embracing_1, echo = TRUE}

calculateMean <- function(data, variable){
  
  data %>% 
    summarize("mean" = mean({{variable}}))
  
}

calculateMean(storm_sample, pressure)

```
This example allows a user to pass an entire dataframe into a function, and randomly sample a number of rows, as provided by the user.

```{r embracing_2, echo = TRUE , eval = FALSE}

mySample <- function(df, n_rows){
  
  {{df}} %>% 
    sample_n(size = {{n_rows}})
}

mySample(storms, 3)

```

```{r tbl_embracing_2}

mySample <- function(df, n_rows){
  
  {{df}} %>% 
    sample_n(size = {{n_rows}})
}

mySample(storms, 3) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```



# Across
According to the [dplyr documentation](https://dplyr.tidyverse.org/reference/across.html), `across()` provides an easy interface to apply the same transformation to multiple columns.

### Apply a calculation function to multiple columns  
Calculating the mean of all Integer columns, grouping by the Storm's category:
```{r across_sample, eval = FALSE}

storms %>% 
  group_by(category) %>% 
  summarize(across(where(is.integer), mean))


```

```{r tbl_across_sample }

storms %>% 
  group_by(category) %>% 
  summarize(across(where(is.integer), mean)) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```


### Transform all columns of a certain datatype to another datatype
Coercing all **factor** columns to **string** datatypes:

```{r across_type_conversion, echo = TRUE}

storms %>% 
  mutate(across(is.factor,as.character)) %>% 
  glimpse()

```

### Running multiple named functions on multiple columns:  
Calculating the minimum, maximum, standard deviation, and mean of columns that start with the letter "L."

```{r across_listed_functions, echo = TRUE, eval = FALSE}

storms %>% 
  group_by(category) %>% 
  summarize(across(starts_with("l"), list(min_val = min,
                                         max_val = max,
                                         std_dev = sd,
                                         avg = mean)))


```
```{r tbl_across_listed_functions}

storms %>% 
  group_by(category) %>% 
  summarize(across(starts_with("l"), list(min_val = min,
                                         max_val = max,
                                         std_dev = sd,
                                         avg = mean))) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)


```

### Row-wise calculations with **c_across()**  

Though uncommon, there may be instances where we need to do calculations on a row-by-row basis. `dplyr` provides a variation of `across()` to accomplish this: `c_across()`.

Starting with a simple transform of the `storms` dataset, where each row will be a year, each column will be the category of storm, and each value represents the average wind speed for the corresponding year and category. The transformed dataset will be called `windy_years`. 
```{r create_windy_years, echo = FALSE}

# Transform the dataset to a wide format, to prepare for a row-wise calculation.

windy_years <- storms %>% 
  group_by(year, category) %>% 
  summarize("avg_wind" = mean(wind, na.rm = TRUE)) %>% 
  ungroup() %>% 
  tidyr::pivot_wider(names_from = category, values_from = avg_wind)

```

For the sake of example, calculate a somewhat meaningless row-wise mean on each row of this table -- assigning the value to `avg`. 

```{r preview_windy_years, echo = TRUE}

windy_years %>% 
  sample_n(10) %>% # sample 10 rows
  rowwise() %>% 
  mutate("avg" = mean(
              c_across(everything()),
              na.rm = TRUE)
         ) %>% 
  kableExtra::kbl() %>% 
  kable_paper("hover", full_width = F)

```

## Wrapping up  

Though I expect to benefit from all of these enhancements, I think tidy selection will have the most immediate impact. Many of my analysis workflows involve transforming large datasets with many columns into easier-to-work-with tables with the columns that are relevant to my analysis. Tidy selection makes it possible to do this with limited hard-coding. I also expect to use across() in place of repetitive code where I calculate measures on multiple columns. Finally, I think I'll be able to make many of my analysis scripts more concise, combining many existing functions into fewer generalized functions -- all thanks to my new friend `{{ curly curly }}` syntax.



